A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-01 latex/main.tex
Updates from Overleaf

Akaishi and Fujiwara \cite{Akaishi2023} have hypothesized that XX.
We agree on XX, but not others.
Could be limitation of the fact that they didn't analyse many samples.

The placement of the following major deletion events
in the ARG aligns well with the phylogenetic placement of
VOC-associated deletion events
suggested in a previous study \cite{Akaishi2023}:
a 3-base deletion 21,991 above the Alpha origin node;
two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
a 9-base deletion at 28,362 above the B.1.1.529 origin node;
a 3-base deletion at 6,513,
a 9-base deletion at 21,987, and
a 3-base deletion at 22,194 above the BA.1 origin node;
a 9-base deletion at 21,633 above the BA.2 origin node.

However, the ARG suggests that two major deletions
(a 6-base deletion at 21,765 and a 1-base deletion at 28,271;
Table~\ref{tab:major_dels}) are recurrent
(i.e., acquired multiple times),
but Akaishi and Fujiwara \cite{Akaishi2023} suggest that
these deletions might have been acquired
before the divergence of certain VOCs,
based on an analysis of a small set of 20 samples.
That study suggests that the 6-base deletion
occurred before the divergence of Alpha and Omicron,
but in the ARG this deletion occurs three times,
above the origin nodes of Alpha, BA.1, and BA.4.
Similarly, that study suggests that the 1-base deletion
occurred before the divergence of Alpha, Delta, and Omicron,
but in the ARG this deletion occurs
above the Alpha origin node and above node 1436808.

There are three contiguous deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node (starting at 11,283) and
the other above the BA.2 origin node (starting at 11,288)
in the ARG (Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up during deletion remapping
as they had four overlapping sites at 11,288-11,292
(which was mapped as the 4-base deletion event).
Furthermore, Akaishi and Fujiwara \cite{Akaishi2023}
suggest that a 9-base deletion at 11,288 occurred
before the divergence of Alpha and Omicron.
In the ARG, however, this deletion occurs
above the Alpha origin node.

We identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG shown in Table~\ref{tab:rec_dels}.
These include known recurrent deletions, for example,
6-base deletion at 21,765,
3-base deletion at 21,991, and
3-base deletion 22,194 in S1-NTD
associated with antibody escape \cite{McCarthy2021,Meng2021}.
The out-of-frame deletions in S1-NTD
(e.g., the 2-base deletion at 21,765),
N, ORF1ab, and ORF8
are implausible and are likely artefacts caused by
sequencing errors or sequence misalignments.
Future improvements in
data pre-processing and deletion remapping
should remove such deletions.

@@

The displayed subgraph shows a small selection of the 192 XB samples.
XB was originally defined as a recombinant between
the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
from North and Central America \cite{Gutierrez2022}.
However, these parent lineages were not added to the ARG due to a high HMM cost.
Together with other lineages such as B.1.627
(sister to B.1.631 \cite{Gutierrez2022}),
they are likely to attach somewhere along the long branch
in the subgraph visualization (with >30 mutations),
which also contains the key "R2" deletion
associated with XB / B.1.634 \cite{Gutierrez2022}.

>>>>

Based on a phylogenetic analysis of 25 randomly chosen samples,
Akaishi and Fujiwara \cite{Akaishi2023} hypothesized
the emergence of commonly observed indels in SARS-CoV-2.
We found support in the ARG for the the phylogenetic placement of
these VOC-associated deletion events suggested in that study:
a 3-base deletion 21,991 above the Alpha origin node;
two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
a 9-base deletion at 28,362 above the B.1.1.529 origin node;
a 3-base deletion at 6,513,
a 9-base deletion at 21,987, and
a 3-base deletion at 22,194 above the BA.1 origin node;
a 9-base deletion at 21,633 above the BA.2 origin node.
However, the ARG suggests that two major deletions
(a 6-base deletion at 21,765 and a 1-base deletion at 28,271;
Table~\ref{tab:major_dels}) are recurrent
(i.e., acquired multiple times),
but Akaishi and Fujiwara \cite{Akaishi2023} suggest that
these deletions might have been acquired
before the divergence of certain VOCs,
based on an analysis of a small set of 20 samples.
That study suggests that the 6-base deletion
occurred before the divergence of Alpha and Omicron,
but in the ARG this deletion occurs three times,
above the origin nodes of Alpha, BA.1, and BA.4.
Similarly, that study suggests that the 1-base deletion
occurred before the divergence of Alpha, Delta, and Omicron,
but in the ARG this deletion occurs
above the Alpha origin node and above node 1436808.

We identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG shown in Table~\ref{tab:rec_dels}.
These include known recurrent deletions, for example,
6-base deletion at 21,765,
3-base deletion at 21,991, and
3-base deletion 22,194 in S1-NTD
associated with antibody escape \cite{McCarthy2021,Meng2021}.

% SHZ: Talk about the agreement before getting into improvements.

There are three contiguous deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node (starting at 11,283) and
the other above the BA.2 origin node (starting at 11,288)
in the ARG (Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up during deletion remapping
as they had four overlapping sites at 11,288-11,292
(which was mapped as the 4-base deletion event).
Furthermore, Akaishi and Fujiwara \cite{Akaishi2023}
suggest that a 9-base deletion at 11,288 occurred
before the divergence of Alpha and Omicron.
In the ARG, however, this deletion occurs
above the Alpha origin node.
The out-of-frame deletions in S1-NTD
(e.g., the 2-base deletion at 21,765),
N, ORF1ab, and ORF8
are implausible and are likely artefacts caused by
sequencing errors or sequence misalignments.
Future improvements in
data pre-processing and deletion remapping
should remove such deletions.

@@

The displayed subgraph shows a small selection of the 192 XB samples.
XB was originally defined as a recombinant between
the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
from North and Central America \cite{Gutierrez2022}.
However, these parent lineages were not added to the ARG due to a high HMM cost.
Together with other lineages such as B.1.627
(sister to B.1.631 \cite{Gutierrez2022}),
they are likely to attach somewhere along the long branch
in the subgraph visualization (with > 30 mutations),
which also contains the key "R2" deletion
associated with XB / B.1.634 \cite{Gutierrez2022}.
